MCTS Based on Simple Regret

### 2012-01-21 paper/rmcts/aaai/paper.tex
future work

We begin with introduction of several sampling schemes with better
bounds for the simple regret on sets. We then extend the resutls to
sampling in trees by combining the proposed sampling schemes on the
first step of a rollout with UCT during the rest of the
rollout. Finally, we empirically evaluate the performance of the
proposed sampling schemes on sets of Bernoulli arms, in randomly
generated trees, and on the sailing domain.

@@

\begin{dfn} Scheme $\mathbf{UCB(\alpha)}$ repeatedly pulls arm $i$ that maximizes 
upper confidence bound $b_i$ on the reward:
\begin{equation}
b_i=\overline X_i+\sqrt {\frac {\alpha \ln n} {n_i}}
\end{equation}
where $\overline X_i$ is the average reward obtained from arm $i$,
$n_i$ is the number of times arm $i$ was pulled, and $n$ is the total
number of pulls so far. \end{dfn} The best known upper bound on the simple
regret of UCB($\alpha$) is polynomially decreasing in the number of samples
(see \cite{Bubeck.pure}, Theorems~2,3).

@@

We introduce here two sampling schemes with exponentially
decreasing upper bounds on the simple regret. The bounds
suggest that these schemes achieve a lower simple regret
than uniform sampling; indeed, this is confirmed
by experiments.

We first consider $\varepsilon$-greedy sampling as a straightforward
generalization of uniform sampling:
\begin{dfn} The \textbf{$\mathbf{\varepsilon}$-greedy} sampling scheme
pulls the empirically best arm with probability $\varepsilon$ and any other
arm with probability $\frac {1-\varepsilon} {K-1}$. 
\end{dfn}

\begin{thm} Simple regret of the  $\varepsilon$-greedy
sampling scheme is asymptotically bounded from above as
\begin{equation}
\IE r_{\varepsilon\mbox{-}greedy}\le...
\end{equation}
the bound is minimized for $\varepsilon=\frac 1 2$.
\end{thm}

\begin{proof}[Proof sketch:] TODO
\end{proof}

\begin{dfn} Scheme $\mathbf{UCB_{\sqrt{\cdot}}(\alpha)}$ repeatedly pulls arm $i$ that
maximizes $b_i$:
\begin{equation}
b_i=\overline X_i+\sqrt {\frac {\alpha \sqrt n} {n_i}}
\end{equation}
where, as before, $\overline X_i$ is the average reward obtained from arm $i$,
$n_i$ is the number of times arm $i$ was pulled, and $n$ is the total
number of pulls so far. \end{dfn}

\begin{thm} Simple regret of the UCB$_{\sqrt{}}$($\alpha$) sampling
  scheme is asymptotically bounded from above as.
\begin{equation}
\IE r_{ucb\sqrt{\cdot}} \le ...
\end{equation}
\end{thm}
\begin{proof}[Proof sketch:] TODO
\end{proof}

@@

UCT \cite{Kocsis.uct} is a generalization of UCB for MCTS.  UCT
applies UCB at each step of a rollout.  We suggest an improvement on
UCT, which combines different sampling schemes on the first step and
during the rest of a rollout:
\begin{dfn}
The \textbf{SR+CR MCTS sampling scheme} selects an action at the
current root node according to a scheme suitable for minimizing 
simple regret (\textbf{SR}), such as $\frac 1 2$-greedy or UCB$_{\sqrt{\cdot}}$, and
then selects actions according to UCB, minimizing cumulative regret (\textbf{CR}).
\end{dfn}

@@

\subsection{VOI-aware Sampling}

A further improvement can be achieved by computing or estimating the
value of information (VOI) of the rollouts and choosing a rollout that
maximizes the VOI. VOI of a rollout can be computed when the sample
distribution of an action is known up to the parameters, such as the
normal distribution with an unknown mean and/or
variance. Alternatively, the value of information can be estimated
from the set of samples, and the need to assume a particular shape of
the distribution can be lifted. In one realization of the latter
approach the VOI of performing an action is estimated as an
\emph{upper bound on the value of perfect information about an arm,
divided by the number of pulls of the arm}.
\begin{eqnarray}
VOI_\alpha&\approx&\frac {\overline X_\beta} {n_\alpha+1}
\exp\left(-2(\overline X_\alpha - \overline X_\beta)^2 n_\alpha\right)\\
VOI_i&\approx&\frac {1-\overline X_\alpha} {n_i+1}
\exp\left(-2(\overline X_\alpha - \overline X_i)^2 n_i\right),\; i\ne\alpha\nonumber\\
\mbox{where }\alpha&=&\arg\max_i \overline X_i\nonumber\\
             \beta&=&\arg\max_{i,\,i\ne\alpha} \overline X_i\nonumber
\end{eqnarray}
Early experiments with this approach demonstrated
significantly lower simple regret on a wide range of problems.

@@

The results were empirically verified on Multi-armed Bandit instances,
on search trees, and on the sailing domain, as defined in
\cite{Kocsis.uct}. In most cases, the experiments showed lower average
simple regret for $\frac 1 2$-greedy an UCB$_{\sqrt{\cdot}}$ than for
UCB on sets, and for the SR+CR scheme than for UCT in trees.

@@

Figure~\ref{fig:sailing-rcq-vs-factor} compares UCB$_{\sqrt{\cdot}}$+UCT with UCT with a
different exploration factor at the root (UCB+UCT). $\alpha$ for the rest of
the steps was chosen to ensure the best performance from earlier
experiments on the domain. Both algorithms exhibit similar dependency,
but as the number of samples grows, UCB$_{\sqrt{\cdot}}$+UCT achieves smaller average
regrets and is less sensitive to the choice of the value for $\alpha$ at
the first step.
\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.45]{rcq-size=6-nsamples=397.pdf}\\
  b. 397 rollouts\\
  \vspace{1em}
  \includegraphics[scale=0.45]{rcq-size=6-nsamples=1585.pdf}\\
  d. 1585 rollouts\\
  \caption{The sailing domain, log vs. sqrt, $6\times 6$ lake}
  \label{fig:sailing-rcq-vs-factor}
\end{figure}

@@

The MTCS SR+CR scheme presented in the paper differs
from UCT at the first step of the rollout, when the `simple' selection
regret is minimized instead of the cumulative regret. Both the
theoretical analysis and the empirical evaluation provide evidence for
better general performance of the proposed scheme.

>>>>

We begin with introduction of several sampling schemes with better
bounds for the simple regret on sets. We then extend the results to
sampling in trees by combining the proposed sampling schemes on the
first step of a rollout with UCT during the rest of the
rollout. Finally, we empirically evaluate the performance of the
proposed sampling schemes on sets of Bernoulli arms, in randomly
generated trees, and on the sailing domain.

@@

\begin{dfn} Scheme $\mathbf{UCB(\alpha)}$ repeatedly pulls arm $i$ that maximizes 
upper confidence bound $b_i$ on the reward:
\begin{equation}
b_i=\overline X_i+\sqrt {\frac {\alpha \ln n} {n_i}}
\label{eqn:ucb}
\end{equation}
where $\overline X_i$ is the average reward obtained from arm $i$,
$n_i$ is the number of times arm $i$ was pulled, and $n$ is the total
number of pulls so far. \end{dfn} The best known upper bound on the simple
regret of UCB($\alpha$) is polynomially decreasing in the number of samples
(see \cite{Bubeck.pure}, Theorems~2,3).

@@

We introduce here two sampling schemes with superpolynomially
decreasing upper bounds on the simple regret. The bounds
suggest that these schemes achieve a lower simple regret
than uniform sampling; indeed, this is confirmed
by experiments.

We first consider $\varepsilon$-greedy sampling as a straightforward
generalization of uniform sampling:
\begin{dfn} The \textbf{$\mathbf{\varepsilon}$-greedy} sampling scheme
pulls the empirically best arm with probability $\varepsilon$ and any other
arm with probability $\frac {1-\varepsilon} {K-1}$. 
\end{dfn}
This sampling scheme exhibits an exponentially decreasing simple regret:
\begin{thm} For any $\pi>0$  and $\gamma>0$ there exists $N$ such that for
  any number of samples $n>N$ the simple regret of the  $\varepsilon$-greedy
sampling scheme is  bounded from above as
\begin{equation}
\IE r_{\varepsilon\mbox{-}greedy}\le (1+\gamma) \sum_{i=1}^K\Delta_i\exp(\cdot)
\end{equation}
with probability at least $1-\pi$.
the bound is minimized for $\varepsilon=\frac 1 2$.
\end{thm}

\begin{proof}[Proof outline:] TODO
\end{proof}

$\varepsilon$-greedy is based solely on sampling the empirically best
arm with a higher probability then the rest of the arms, and ignores
information about empirical means of other arms. On the other hand,
UCB distributes samples in accordance with empirical means, but, in order to
minimize cumulative regret, chooses the empirically best arm too often.
Intuitively, a better scheme for simple regret minimization would
distribute samples in a way similar to UCB, but would sample the best arm
less. This can be achieved by replacing $\log(\cdot)$ in
Equation~\ref{eqn:ucb} with a faster growing sublinear function, for
example, $\sqrt\cdot$.
\begin{dfn} Scheme $\mathbf{UCB_{\sqrt{\cdot}}(\alpha)}$ repeatedly pulls arm $i$ that
maximizes $b_i$:
\begin{equation}
b_i=\overline X_i+\sqrt {\frac {\alpha \sqrt n} {n_i}}
\end{equation}
where, as before, $\overline X_i$ is the average reward obtained from arm $i$,
$n_i$ is the number of times arm $i$ was pulled, and $n$ is the total
number of pulls so far. \end{dfn}
This scheme also exhibits a superpolynomially decreasing simple regret:
\begin{thm}  For any $\pi>0$  and $\gamma>0$ there exists $N$ such that for
  any number of samples $n>N$ the simple regret of the  UCB$_{\sqrt{}}$($\alpha$)
sampling scheme is  bounded from above as
\begin{equation}
\IE r_{ucb\sqrt{\cdot}} \le (1+\gamma)\sum_{i=1}^K\Delta_i\exp\left(-2\alpha\sqrt{n}\right)
\end{equation}
with probability at least $1-\pi$.
\end{thm}

\begin{proof}[Proof outline:] TODO
\end{proof}

@@

UCT \cite{Kocsis.uct} is a generalization of UCB for MCTS.  UCT
applies UCB at each step of a rollout.  We suggest an improvement on
UCT, which combines different sampling schemes on the first step and
during the rest of a rollout:
\begin{dfn}
The \textbf{SR+CR MCTS sampling scheme} selects an action at the
current root node according to a scheme suitable for minimizing 
the simple regret (\textbf{SR}), such as $\frac 1 2$-greedy or UCB$_{\sqrt{\cdot}}$, and
then selects actions according to UCB, minimizing the cumulative regret (\textbf{CR}).
\end{dfn}

@@

\subsection{VOI-aware Sampling}
\label{sec:voi-sampling}

A further improvement can be achieved by computing or estimating the
value of information (VOI) of the rollouts and choosing a rollout that
maximizes the VOI. VOI of a rollout can be computed when the sample
distribution of an action is known up to the parameters, such as the
normal distribution with an unknown mean and/or
variance. Alternatively, the value of information can be estimated
from the set of samples, and the need to assume a particular shape of
the distribution can be lifted. In one realization of the latter
approach the VOI of performing an action is estimated as an
\emph{upper bound on the value of perfect information about an arm,
divided by the number of pulls of the arm}.
\begin{eqnarray}
VOI_\alpha&\approx&\frac {\overline X_\beta} {n_\alpha+1}
\exp\left(-2(\overline X_\alpha - \overline X_\beta)^2 n_\alpha\right)\\
VOI_i&\approx&\frac {1-\overline X_\alpha} {n_i+1}
\exp\left(-2(\overline X_\alpha - \overline X_i)^2 n_i\right),\; i\ne\alpha\nonumber\\
\mbox{where }\alpha&=&\arg\max_i \overline X_i\nonumber\\
             \beta&=&\arg\max_{i,\,i\ne\alpha} \overline X_i\nonumber
\end{eqnarray}
Early experiments with this approach demonstrated
a significantly lower simple regret on a wide range of problems.

@@

The results were empirically verified on Multi-armed Bandit instances,
on search trees, and on the sailing domain, as defined in
\cite{Kocsis.uct}. In most cases, the experiments showed a lower average
simple regret for $\frac 1 2$-greedy an UCB$_{\sqrt{\cdot}}$ than for
UCB on sets, and for the SR+CR scheme than for UCT in trees.

@@

Figure~\ref{fig:sailing-rcq-vs-factor} compares UCB$_{\sqrt{\cdot}}$+UCT with UCT with a
different exploration factor at the root (UCB+UCT). $\alpha$ for the rest of
the steps was chosen to ensure the best performance from earlier
experiments on the domain. Both algorithms exhibit similar dependency,
but as the number of samples grows, UCB$_{\sqrt{\cdot}}$+UCT achieves smaller average
regrets and is less sensitive to the choice of the value for $\alpha$ at
the first step.
\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.45]{rcq-size=6-nsamples=397.pdf}\\
  b. 397 rollouts\\
  \vspace{1em}
  \includegraphics[scale=0.45]{rcq-size=6-nsamples=1585.pdf}\\
  d. 1585 rollouts\\
  \caption{The sailing domain, UCB vs. UCB$_{\sqrt{\cdot}}$, $6\times 6$ lake}
  \label{fig:sailing-rcq-vs-factor}
\end{figure}

@@

The MCTS SR+CR scheme presented in the paper differs
from UCT at the first step of the rollout, when the `simple' selection
regret is minimized instead of the cumulative regret. Both the
theoretical analysis and the empirical evaluation provide evidence for
better general performance of the proposed scheme.

The improvement is inspired by the notion of value of information
(VOI), but VOI is used implicitly in the analysis of the algorithm,
rather than computed or learned explicitly in order to plan the
rollouts. Preliminary results on VOI-aware sampling suggest that a
further improvement in the sampling performance can be
achieved. However, application of the theory of rational metareasoning
to Monte Carlo Tree Search is an open problem \cite{HayRussell.MCTS},
and both a solid theoretical model and empirically efficient VOI
estimates still need to be developed.
